Browsernative Microservices

### 2017-02-27 build/paper.tex
pdf/tex build

\begin{document}
\maketitle
\begin{abstract}
Building complex web applications nowadays require additional layers of
abstraction and often heavily depend on proprietary frameworks. New
specifications build right into the browserengine provide a native
service API to overcome tricky abstraction constraints.
\end{abstract}

@@

Derived from the Latin word \textbf{simplex}, simple can be understood
as ``literally, uncompounded or onefold''\footnote{\href{http://www.etymonline.com/index.php?term=simple}{Etymology
  Dictionary}} which points towards an unidimensional nature. While
complexity describes the multilayered und entangled nature of
conditions, simplicity empowers the human brain to reason about issues
in a straightforward manner. It certainly has some overlapping's with
easy, but while easy is more of a relative spirit, simple can be laid
out as a objective manner and therefore universally applicable.

@@

In the last four years the average transfer size of a webpage doubled to
currently around 2.5 MB.\footnote{\href{http://httparchive.org/trends.php}{HTTPArchive
  Trends}} Leaving images, fonts or other content aside the size of
HTML, CSS and JS sums up to a total average of 550 kb. One character
weights around 1 byte which means an average webpage is delivering
550.000 character or around 125 pages of single-spaced text. Frederic
Filloux analyzed the payload on different newspaper websites and came to
the conclusion, that only round about 5-6 \% of the transferred
characters made for human consumption.{[}1{]}

@@

In the recent years many \textbf{frameworks}, libraries and
methodologies approached the global nature of the DOM by scoping assets
and design rules into maintainable components. While the DOM can't be
scoped, JS can. Many frameworks, like ReactJS, AngularJS or VueJS just
to name a few, ditched the old rule of separated HTML, CSS and JS in
favor of an additional layer of abstracted JS components (containing
structure, behavior and sometimes styling). Quiet often those frameworks
mimic a MVC pattern on top of the browser engine which is a reasonable
simple design pattern to build graphical user interfaces. While
frameworks are a valid approach for building scalable web applications
they remain highly opinionated, embody inherent complexity themselves
and can change and break over time. Another downside is code inflation
which is a crucial point for performance. All of those bottlenecks in
the web demand for a new standardized way for creating and evolving
complex web services.

@@

The first principle of services is the \textbf{loose coupling
principle}: changing and deploying one service shouldn't result in
changing other parts of the system.{[}3, p. 30{]}. Mutations or shadowed
variables, which is happening a lot in CSS, making it hard to keep
changes ought to only affect one place in the application. A
\emph{browsernative microservice} therefore pushing encapsulation and
avoiding variable mutations outside its scope as much as possible.
Practically, CSS will be scoped and JS fosters immutable JS entities and
avoids variables leaking into the global namespace.

The second principle of services is the \textbf{high cohesion
principle}: Whether designing a microservice or it's components we want
related behavior sit together, and unrelated behavior to sit
elsewhere.{[}3, p. 30{]} High cohesion can be enhanced towards the more
dynamic \textbf{Single Responsibility Principle}: ``Gather together
those things that change for the same reason and separate those things
that change for different reasons.''{[}5{]} In a very quick and dirty
code quality analysis, the quality can be measured just by counting the
places changes in the code occur in order to implement a functionality.
An arbitrary threefold MVC system should require a maximum of three
changes to implement or change functionalities. The problem in browser
based development is not only the global paradigm which makes changes
deliberately unpredictable. The high cohesion principle is violated by
the traditional separation along the siloed entities \textbf{HTML, JS
and CSS}. Understanding the relation of HTML markup and another CSS file
adds incidental complexity. Different approaches emerged over the recent
years to join the forces. Most notable the dedicated frontend language
\emph{elm}\footnote{\href{http://elm-lang.org/}{elm lang}} where MODEL,
VIEW and CONTROLLER sit in one place. A browsernative service aims for a
combinement of the web native trinity HTML, JS and CSS.

Traditional web development relies on \textbf{libraries} to enhance the
service capabilities of the web platform. Compared to libraries, a
component service offers multiple advantages for building, deployment
and shipping. As expressed earlier a \emph{components for the web}, or
web component, is \textbf{self-contained} which means it embodies all
needed functionality to get it's job done. Therefore it has a much
better evolution mechanism in the service contracts. Changing
functionality won't break other services. A component can progressively
enhanced which guarantees functionality throughout different versions. A
library is only loosely coupled to the implementation and therefore hard
to track in functionality. Changing a library may result in an
unforeseen amount of time fixing implementations. It is not unusual to
see websites embodying different versions of the same library to
guarantee functionality but lowers page performance significantly.

@@

After getting confidence in microservice principles and technical
background the paper should briefly join them to form a
\emph{browsernative microservices}. Needless to say the following
example is overall simplified towards illustrating the connection
between browsernative technologies and microservice patterns. Googles
library Polymer is a good place for learning about web components in
depth and make use of their toolbox. One of their most famous proof of
concept is the so-called
\href{https://shop.polymer-project.org/}{Polymer Shop} which is a
fully-fledged online shop nested within a single root element
\texttt{\textless{}shop-app\textgreater{}}. This app made of several
main views and many more invisible wrapper elements for routing, service
worker caching, theming, etc. The whole shop runs as a single
application fetching and updating remote resources and switching views.
Let's assume we work in a sales engineering team of the Polymer Shop and
need to rebuild the checkout microservice.

@@

Lets start the service description top-down beginning with the
\textbf{root container} managing the overall service.

\subsection{Root container}\label{root-container}

@@

\subsection{Container components}\label{container-components}

The second layer of the checkout microservice like
\texttt{\textless{}sign-in\textgreater{}} or
\texttt{\textless{}shipping-details\textgreater{}} still contain a lot
of logic and no or less styling. Their job is to control their
underlying presentational leaf components, aggregating events and to
define a \textbf{set of actions towards the model} . Every container
mounted directly under the service root may have a dedicated area inside
the worker thread where it's fulfil his duties. For example, a typical
\texttt{\textless{}sign-in\textgreater{}} contains merely two fields
username and password and a submit button. The component listens for the
submit action, aggregating the credentials and might add some semantics
to it like \emph{action: `SIGNIN SUBMITTED'}. Fields and action message
will be dispatched towards the model for further processing like
initiating a authorization process.

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class} \NormalTok{SimpleContainer }\KeywordTok{extends} \NormalTok{HTMLElement }\OperatorTok{\{}
  \CommentTok{// constructor, static methods ...}
  \NormalTok{set }\AttributeTok{_dispatch}\NormalTok{(msg) }\OperatorTok{\{}
    \KeywordTok{this}\NormalTok{.}\AttributeTok{dispatchEvent}\NormalTok{(}
      \KeywordTok{new} \AttributeTok{CustomEvent}\NormalTok{(}\StringTok{'checkout'}\OperatorTok{,} \OperatorTok{\{}
      \DataTypeTok{bubbles}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}
      \DataTypeTok{detail}\OperatorTok{:} \NormalTok{msg}
      \OperatorTok{\}}\NormalTok{)}
    \NormalTok{)}\OperatorTok{;}
  \OperatorTok{\}}
  \NormalTok{set }\AttributeTok{_receive}\NormalTok{(msg) }\OperatorTok{\{}
  \CommentTok{// switch to methods}
    \VariableTok{console}\NormalTok{.}\AttributeTok{log}\NormalTok{(}\KeywordTok{this}\NormalTok{.}\AttributeTok{localName} \OperatorTok{+} \StringTok{"received msg"}\NormalTok{)}\OperatorTok{;}
    \VariableTok{console}\NormalTok{.}\AttributeTok{log}\NormalTok{(msg)}\OperatorTok{;}
  \OperatorTok{\}}
\OperatorTok{\}}
\end{Highlighting}
\end{Shaded}

@@

\subsection{Presentational components}\label{presentational-components}

In comparison to the former presentational component is build mostly

template switch

css

shadow dom

@@

Following a \textbf{browsernative first} principle this paper didn't
made use of third-party libraries so far. Contrary to the example in the
last section there are a few reasons why web components shouldn't be
used without any proprietary technology. At first, from a technical
perspective web components are relatively young and at the time of
writing only the newer Chrome and Opera browsers can handle them
correctly. Most probably it will take a long time until native web
components can be safely used without additional legacy support.The
second inevitable question is the notion if they should be used pure
from a \textbf{simplicity perspective} without any proprietary
technology now and future.

Talking about web components it is impossible to avoid technologies like
\textbf{React} altogether. In the last couple years React introduced
some good concepts to web development like the notion of state, passing
data as properties and declarative event management just to name a few.
Many of those design decisions were made to overall \textbf{simplify
browser development through standardized methodologies}. The principles
of microservices could be applied to React development maybe even
further as for example the communication model involves a lot of
delegation and decomposition.

Many of those features and design decisions in React created much of a
hype around the library and technologies like the virtual DOM while web
components never caught up in momentum at a comparable rate. According
to Github Googles Polymer project is even six months older than React
but has less than one third of its stars. Other notable web components
projects like Bosonic\footnote{https://github.com/bosonic/bosonic/} or
Mozilla's X-tag\footnote{https://github.com/x-tag/core} are more dead
than alive, too.

It seems like the web developing community in a broader sense demands
for features and standardization in frameworks like React (and many
others to be fair). Web components and related libraries clearly missed
a selling point in providing standardized ways for developing scalable
components and applications or being to cumbersome to use at all. For
the future it remains unclear if web components will eventually gain
wider adoption within the UI developing community or will be adopted as
low-level technology for framework development emphasized by Sebastian
Markbage, one of the React creators.{[}18{]}

One interesting approach for building web components is the ``reactish''
Skate\footnote{https://github.com/skatejs/skatejs/}. Following the
functional rendering model from React it combines native methods with
additional functionalities known from React like event management. Given
the focus on native technologies it only weights around 4kb (minified
and gzipped) which is ten times less than React. \textbf{Small
encapsulated micro UI frameworks} like Skate could be certainly the near
future of web development and will hopefully gain more attention. The
more mature UI framework Riot already expressed their intention to
gradually develop towards web components.\footnote{http://riotjs.com/compare/\#web-components}

>>>>

\begin{document}
\maketitle
\begin{abstract}
Building complex web applications nowadays require additional layers of
abstraction and often heavily depend on indispensable frameworks. While
they perfectly circumstance the global paradigm of the DOM they remain
highly proprietary attached to the framework. New w3c specifications
build right into the browser provide a native service API in the need to
create standardized web components for the browser. As of 2017 all major
browsers will ship those technologies in their browsers. The paper not
only provides an introduction to the specification. As an assessment
criteria of it will match those specifications against the concept of
microservices which is an modular system architecture to build scalable
applications.
\end{abstract}

@@

Derived from the Latin word \textbf{simplex} simple can be understood as
``literally, uncompounded or onefold''\footnote{\href{http://www.etymonline.com/index.php?term=simple}{Etymology
  Dictionary}} which points towards an unidimensional state. While
complexity describes the multilayered und entangled nature of conditions
simplicity empowers the human brain to reason about issues
straightforward. It certainly has some overlapping's with easy, but
while easy is more of a relative spirit, simple can be laid out as a
objective manner and therefore universally applicable.

@@

In the last four years the average transfer size of a webpage doubled to
currently around 2.5 MB.\footnote{\href{http://httparchive.org/trends.php}{HTTPArchive
  Trends}} Leaving images, fonts or other content aside the size of
HTML, CSS and JS sums up to an total average of 550 KB. One character
weights around 1 byte which means an average webpage is delivering
550.000 character or around 125 pages of single-spaced text. Frederic
Filloux analyzed the payload on different newspaper websites and came to
the conclusion, that only round about 5-6 \% of the transferred
characters made for human consumption.{[}1{]}

@@

In the recent years many \textbf{frameworks}, libraries and
methodologies approached the global nature of the DOM by scoping assets
and design rules into maintainable components. While the DOM can't be
scoped, JS can. Many frameworks, like React, Angular or Vue just to name
a few, ditched the old rule of separated HTML, CSS and JS in favor of an
additional layer of abstracted JS components (containing structure,
behavior and sometimes styling). Quiet often those frameworks mimic a
MVC pattern on top of the browser engine which is a reasonable simple
design pattern to build UIs. While frameworks are a valid approach for
building scalable web applications they remain highly opinionated embody
inherent complexity themselves and can change and break over time.
Another downside is code inflation which is a crucial point for
performance. All of those bottlenecks in the web demand for a new
standardized way for creating and evolving complex web services.

@@

The first principle of services is the \textbf{loose coupling
principle}: changing and deploying one service shouldn't result in
changing other parts of the system.{[}3, p. 30{]} Picking CSS for
example, where variables tend to be shadowed by higher specific values
making it increasingly difficult to keep changes ought to only affect
one place in the application. A \emph{browsernative microservice}
therefore pushing encapsulation and avoiding variable mutations outside
its scope as much as possible. One solution would be scoped CSS / JS to
avoid variables leaking into the global namespace as well as hidden
implementation details to avoid mutating values.

The second principle of services is the \textbf{high cohesion
principle}: Whether designing a microservice or it's components we want
related behavior sit together and unrelated behavior to sit
elsewhere.{[}3, p. 30{]} High cohesion can be enhanced towards the more
dynamic \textbf{Single Responsibility Principle}: ``Gather together
those things that change for the same reason and separate those things
that change for different reasons.''{[}5{]} In a very quick and dirty
code quality analysis, the quality can be measured just by counting the
places changes in the code occur in order to implement a functionality.
An arbitrary threefold MVC system should require a maximum of three
changes to implement or change functionalities. The problem in browser
based development is not only the global paradigm which makes changes
deliberately unpredictable. The high cohesion principle is violated by
the traditional separation along the siloed entities \textbf{HTML, JS
and CSS}. Understanding the relation of HTML markup and another CSS file
adds incidental complexity. Different approaches emerged over the recent
years to join the forces. Most notable the dedicated frontend language
\emph{elm}\footnote{\href{http://elm-lang.org/}{elm lang}} where MODEL,
VIEW and CONTROLLER sit in one place. A \emph{browsernative service}
sought for an combination of the web native trinity HTML, JS and CSS.

Traditional web development relies on \textbf{libraries} to enhance the
service capabilities of the web platform. Compared to libraries a
component service offers multiple advantages for building, deployment
and shipping. As expressed earlier a \emph{components for the web}, or
web component, is \textbf{self-contained} which means it embodies all
needed functionality to get it's job done. Therefore it has a much
better evolution mechanism in the service contracts. Changing
functionality won't break other services. A component can progressively
enhanced which guarantees functionality throughout different versions. A
library is only loosely coupled to the implementation and therefore hard
to track in functionality. Changing a library may result in an
unforeseen amount of time fixing implementations. It is not unusual to
see websites embodying different versions of the same library to
guarantee functionality but lowers page performance significantly.

@@

After getting confidence in microservice principles and technical
background the paper should briefly join them to form a
\emph{browsernative microservices}. Needless to say the following
example is overall simplified towards illustrating the connection
between browsernative technologies and microservice patterns.
Furthermore it is highly opinionated in development and shouldn't
perceived as ``single source of truth''.

Googles library Polymer is a good place for learning about web
components in depth and make use of their toolbox. One of their most
famous proof of concept is the so-called
\href{https://shop.polymer-project.org/}{Polymer Shop} which is a
fully-fledged online shop nested within a single root element
\texttt{\textless{}shop-app\textgreater{}}. This app made of several
main views and many more invisible wrapper elements for routing, service
worker caching, theming, etc. The whole shop runs as a single
application fetching and updating remote resources and switching views.
Let's assume we work in a sales engineering team of the Polymer Shop and
need to rebuild the checkout microservice.

@@

Lets start the service description top-down beginning with the
\textbf{service root container} managing the overall service.

\subsection{4.1. Service root}\label{service-root}

@@

\subsection{4.2. Container components}\label{container-components}

The second layer of the checkout microservice like
\texttt{\textless{}sign-in\textgreater{}} or
\texttt{\textless{}shipping-details\textgreater{}} still contain mostly
logic and no or less styling. Their job is to control their underlying
presentational leaf components, aggregating events and to define a
\textbf{set of actions towards the model} . Every container mounted
directly under the service root may have a dedicated area inside the
worker thread where it's fulfil his duties. For example, a typical
\texttt{\textless{}sign-in\textgreater{}} contains merely two fields
username and password and a submit button. The component listens for the
submit action, aggregating the credentials and might add some semantics
to it like \emph{action: `SIGNIN SUBMITTED'}. Fields and action message
will be dispatched towards the model for further processing like
initiating a authorization process.

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class} \NormalTok{SimpleContainer }\KeywordTok{extends} \NormalTok{HTMLElement }\OperatorTok{\{}
  \CommentTok{// constructor, static methods ...}
  \NormalTok{set }\AttributeTok{_dispatch}\NormalTok{(msg) }\OperatorTok{\{}
    \KeywordTok{this}\NormalTok{.}\AttributeTok{dispatchEvent}\NormalTok{(}
      \KeywordTok{new} \AttributeTok{CustomEvent}\NormalTok{(}\StringTok{'checkout'}\OperatorTok{,} \OperatorTok{\{}
      \DataTypeTok{bubbles}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}
      \DataTypeTok{detail}\OperatorTok{:} \NormalTok{msg}
      \OperatorTok{\}}\NormalTok{)}
    \NormalTok{)}\OperatorTok{;}
  \OperatorTok{\}}
  \NormalTok{set }\AttributeTok{_receive}\NormalTok{(msg) }\OperatorTok{\{}
    \CommentTok{// switch to methods}
    \VariableTok{console}\NormalTok{.}\AttributeTok{log}\NormalTok{(}\VerbatimStringTok{`}\SpecialCharTok{$\{}\KeywordTok{this}\NormalTok{.}\AttributeTok{localName}\SpecialCharTok{\}}\VerbatimStringTok{ received }\SpecialCharTok{$\{}\NormalTok{msg}\SpecialCharTok{\}}\VerbatimStringTok{`}\NormalTok{)}\OperatorTok{;}
  \OperatorTok{\}}
\OperatorTok{\}}
\end{Highlighting}
\end{Shaded}

@@

\subsection{4.3. Presentational
components}\label{presentational-components}

In comparison to the former container the presentational component is
build mostly around views and styling containing less or no logic
towards handling the component. To extend the last
\texttt{\textless{}sign-in\textgreater{}} container example a typical
presentational component structure could look like the following
top-level structure

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{<sign-in-container>}
  \KeywordTok{<sign-in-presentational>}
  \KeywordTok{</sign-in-presentational>}
\KeywordTok{</sign-in-container>}
\end{Highlighting}
\end{Shaded}

The \texttt{\textless{}sign-in-presentational\textgreater{}} contains
all required input fields and oauth connectors to Google or Facebook but
\textbf{will not care about events they create}. A presentation
component is usually steered via HTML attributes. Contrary to the former
containers the presentational component highly utilizes \textbf{shadow
DOM and HTML templates}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{> sign-in-presentational.html}
\CommentTok{<!-- IMPORTS -->}
\KeywordTok{<link}\OtherTok{ rel=}\StringTok{"import"}\OtherTok{ href=}\StringTok{"facebook-oauth.html"}\OtherTok{ async}\KeywordTok{>}
\KeywordTok{<link}\OtherTok{ rel=}\StringTok{"import"}\OtherTok{ href=}\StringTok{"high-res-template.html"}\OtherTok{ async}\KeywordTok{>}
\NormalTok{...}

\CommentTok{<!-- VIEW -->}
\KeywordTok{<sign-in-presentational>}
  \KeywordTok{<high-res-template>}
    \KeywordTok{<facebook-oauth></facebook-oauth>}
    \KeywordTok{<slot}\OtherTok{ name=}\StringTok{"form"}\KeywordTok{></slot>}
    \KeywordTok{<slot}\OtherTok{ name=}\StringTok{"button"}\KeywordTok{></slot>}
  \KeywordTok{</high-res-template>}

  \KeywordTok{<template}\OtherTok{ id=}\StringTok{"lowres"}\KeywordTok{>}\NormalTok{...}\KeywordTok{</template>}
  \KeywordTok{<template}\OtherTok{ id=}\StringTok{"mobile"}\KeywordTok{>}\NormalTok{...}\KeywordTok{</template>}
\KeywordTok{</sign-in-presentational>}

\KeywordTok{<script>}
  \KeywordTok{class} \NormalTok{SignIn }\KeywordTok{extends} \NormalTok{HTMLElement }\OperatorTok{\{}
    \AttributeTok{constructor}\NormalTok{() }\OperatorTok{\{}
      \KeywordTok{super}\NormalTok{()}\OperatorTok{;}
      \KeywordTok{this}\NormalTok{.}\AttributeTok{attachShadow}\NormalTok{(}\OperatorTok{\{}\DataTypeTok{mode}\OperatorTok{:} \StringTok{'open'}\OperatorTok{\}}\NormalTok{)}\OperatorTok{;}
      \CommentTok{// append templates according to window.innerWidth}
      \CommentTok{// run this.getAttribute('htmlprop') to control element}
    \OperatorTok{\}}
    \AttributeTok{attributeChangedCallback}\NormalTok{(attrName}\OperatorTok{,} \NormalTok{oldVal}\OperatorTok{,} \NormalTok{newVal) }\OperatorTok{\{}
      \CommentTok{// react on changing attributes}
    \OperatorTok{\}}
  \OperatorTok{\}}
  \VariableTok{customElements}\NormalTok{.}\AttributeTok{define}\NormalTok{(}\StringTok{'sign-in-presentational'}\OperatorTok{,} \NormalTok{SignIn)}\OperatorTok{;}
\OperatorTok{<}\SpecialStringTok{/script>}
\end{Highlighting}
\end{Shaded}

Presentational components require configuration and usually ship a long
list of CSS selectors. Whereas it's possible to expand templates in
place it could be also possible to define \textbf{template components}
to atomize the component further and increase code readability.

tests

@@

Following a \textbf{browsernative first} dogma this paper didn't put
third-party libraries into view so far in the search of an independent
approach. Nevertheless, it should be clear to this point that the
example in the last section is far away from being usable in production.
At first, from a technical perspective web components are relatively
young and at the time of writing only usable in \textbf{newer Chrome and
Opera browsers}. Most probably it will take a long time until native web
components can be safely used without additional legacy support. The
troublesome \emph{HTML imports} spec will certainly need even longer
which inevitable opens up the question if web components are useful from
today on. And second, do web components align with the notion of
simplicity if too many downsides and possible pitfalls require
attention?!

Any profound discussion about web components would be incomplete without
touching component alternatives like \textbf{React}. In the last couple
years React introduced new concepts to web development like the idea of
state, passing data as properties and declarative event management just
to name a few. Those design decisions were made to overall simplify
browser development with full browser support right from the start.
Overall, Reacts perception about UI development created much of a hype
around the library and technologies like the virtual DOM while web
components never caught up in momentum at a comparable rate. According
to Github Googles Polymer project is even six months older than React
but has less than one third of its stars. Other notable web components
projects like Bosonic\footnote{https://github.com/bosonic/bosonic/} or
Mozilla's X-tag\footnote{https://github.com/x-tag/core} are more dead
than alive, too.

It seems like the web developing community in a broader sense values
\textbf{standardized procedures, tooling and browser support} in
frameworks like React (and many others to be fair). Web components and
related libraries clearly missed a real selling point towards React and
Angular. Providing guidelines, simpler syntax and features missed out or
being to cumbersome to use at all. For the future it remains unclear if
web components will eventually gain wider adoption within the UI
developing community or will be adopted as low-level technology for
framework development emphasized by Sebastian Markbage, one of the React
creators.{[}18{]}

Even though the downsides seems heavy weight there are notable attempts
to bring web components into production. A very good example is the
``reactish'' library Skate\footnote{https://github.com/skatejs/skatejs/}.
Following the functional rendering model from React it combines native
methods with additional functionalities known from React like
declarative event management. Given the focus on native technologies it
only weights around 4kb (minified and gzipped) which is ten times less
than React. \textbf{Small encapsulated micro UI frameworks} like Skate
could certainly be the near future of web development and will hopefully
pave the way of web components.

The role of web components may not result in an isolated framework but
move towards a common denominator for future Reacts, Angulars,
Vues\ldots{} which makes it worth considering it as recyclable
technology foundation. The more mature UI framework Riot already
expressed their intention to gradually develop towards web
components.\footnote{http://riotjs.com/compare/\#web-components}
